Are populations like a circuit? Comparing isolation by resistance to a new coalescent-based method.

### 2019-04-17 review-responses.tex

    Overall, I think this article makes a very valid contribution for Molecular
    Ecology Ressources. The authors put their work into the right context and
    show good knowledge of the relevant literature. The writing is both clear
    and concise. Importantly, they do not shy away from pointing out the limits
    of their method, and the very stringent (often unrealistic) assumptions
    such methods usually make.

    I have several comments, but none of them are ``make-or-break'' or fatal flaws.
    The manuscript is already in a good state. But I hope the authors can use my
    comments to further improve their work.  
\end{quote}

\begin{point}{}
    Clearer point out the current computational limitations of your coalescence
time method. It seems it gets tricky with more than about 30 demes? What is the
run-time then, does the whole thing blow up? Give an idea how big is too big!
\end{point}

\reply{
    The runs with 16 demes take around 10 hours with the number of iterations we used
    (noted in the text, \revref).
    Thirty demes is probably a good limit for the current implementation. 
    There are changes that could be made to make it more efficient for larger systems, 
    but at that point the ill conditioning problem would be more severe 
    because there would be more parameters.
    We've added a note about this to the discussion. \llname{ll:computation}
}

@@

\begin{point}{}
    What is the exact scaling? The bottleneck seems to be calculating the
    coalescence times with equation (7). You only say that it ``can be solved
    with standard linear algebra''. Are you doing Gaussian elimination or
    something else worse than $N^2$? Do you use any speed-ups to
    deal with the sparse nature of G for neighbor-only migration? Is it faster
    than getting the coalescence times by iterating time-discretized
    recursions?  It would be helpful for further progress to describe the exact
    scaling and the exact algorithm you use.
\end{point}

\reply{
    In the current implementation, 
    equation (7) is solved using \texttt{solve(A,b)} in R. 
    The actual method varies depending on the specific linear algebra library used, 
    but it is typically Gaussian elimination or LU decomposition, 
    both of which are $\mathcal{O}(N^3)$ 
    (unfortunately the decomposition is different in each step;
    if it were the same it would be $\mathcal{O}(N^2)$).
    For fine discretizations of space, the matrices are sparse,
    but in our testing, using sparse matrix methods didn't help at this scale.
    We've added a point about this last thing to the discussion \revref
    but haven't put in more detail (e.g., plots of scaling with $N$)
    because it depends on the details (sparsity of the adjacency matrix),
    and as we discuss above, we think that nonidentifiability is a more serious barrier
    to finer resolutions.
}

@@

\begin{point}{}
    You could write an outlook stating that your method could be used to fit
    parameterized models with few parameters. For instance, in case the putative
    location of a barrier or an area of biased gene flow is known, one could try to
    infer it without fitting ALL movement parameters, but a simple parametric
    model, which would implicitly take care of the regularization.  
\end{point}

\reply{
    This is a good thing to emphasize. We've added a bit to this effect. \revref
}

\begin{point}{}
    You use mean absolute error in movement rate to describe the model fit. I
    believe this measure can be quite misleading for exponential movement rates.
    Large rates can be notoriously hard to infer (are you super or super-super well
    connected is almost the same), so it is easy to be off quite a bit, which
    drives the average absolute error in rate. Also in practice, one does not care
    so much about absolute rates but ``relative'' ones. So using a correlation
    measure R**2  of inferred movement rates with actual movement rates would be a
    very useful and natural fit parameter.  
\end{point}

\reply{
    This is a good point, and we did experiment with other measures of model fit.
    However, in practice the mean absolute error captured patterns as well as other methods.
    In the interest of not increasing complexity of the paper, we haven't included results
    from the other methods.
}

\begin{point}{}
    You write you use standard methods to check convergence and mixing of the
    MCMC chain \revreffull{1}{6}. Given the degeneracy and the high dimensionality,
    where there any problems? E.g., local optima you were stuck sometimes? Or
    was it always pretty straightforward? Maybe write a few sentences on that!
\end{point}

\reply{
    Convergence could be slow, and the likelihood surface is often ridgy, 
    but we did not see too many problems with getting stuck in completely wrong
    areas of state space.
    This the reason we introduced the non-standard ``pre-burn-in'' phase.
}

\begin{point}{\revref}
    $F_ST \neq$ Absolute divergence!!!  $F_ST$ is more a
    co-variance measure, and for instance heavily influence by sampling a batch
    of related individuals (whereas the mean of absolute divergence will be
    unchanged).
\end{point}

\reply{
    We can't agree more!!!!
    We've added a note about this, although not a lot of detail,
    because it seems outside the scope of this paper. \revref
    And, we thought we should mention $F_{ST}$ because it is commonly used
    in these situations.
}

\begin{point}{}
    You have your own implementation of resistance distance methods, by first
    extracting hitting times. However, you allow non-symmetric rates, something
    no existing resistance method would do. Given your observation that there
    is a lot degeneracy for resistance times (adding skew stuff to hitting
    times), this introduces uncertainty that likely cannot be broken such as
    with coalescence methods. This could drive the high uncertainty in Fig.\ 5
    upper right. Given your good results for Fig.\ 4 for the symmetric case,
    this is a bit surprising. Maybe run one scheme where you FIX to a single
    symmetric movement rate between each pair of demes. This setting would be a
    fairer comparison (and be much closer to what EEMS etc.\ actually do -- they
    do not even try to infer asymmetric rates).
\end{point}

\reply{
    We considered doing this, but
    decided to allow asymmetric migration for resistance distance
    in order to make the two methods more directly comparable.
    For instance, how would we compare to the truth if the truth is asymmetric?
    Comparing methods with both restricted to be symmetric 
    certainly could be useful and informative,
    but we chose not to include it because it seems not to be the most important comparison
    (as reality is probably often asymmetric)
    and we didn't want the paper to get overly complex.
}

@@

\reply{
    We've added a bit more to our big-picture overview that we hope achieves
    the goal of setting the stage appropriately. \revref
}

\begin{point}{\revref}
    Do you actually use this tensor product for the movement
    rates anywhere? And seems Eq.\ 7 (which has no tensor product) is all you
    need.
\end{point}

\reply{
    We included this because we think it's a nice compact way to write, code, 
    and think about the equation,
    and the tensor product equation is actually the one used in the R code.
    It is correct that the equations are equivalent.
}

\begin{point}{Line 115:}
    Not all $ZC$ skew-symmetric would translate to valid generators 
    $G + Z$. The diagonal elements of G have another condition (they have to be minus
    the sum of all other jumping rates). Would be interesting to explore this a
    bit more -- overall what the dimensionality this degeneracy is (implicit
    function theorem), and also if there are simple toy examples where they pop
    up.
\end{point}

\reply{
    Thank you for catching this.
    You are correct that in addition to $ZC$ being skew-symmetric,
    the rows of $Z$ must also sum to zero, which we now say. \revref
    Further exploration of the constraints would be interesting,
    but we think is beyond the scope of this paper.
}

@@

\begin{point}{}
    In the figures, probably keep model figure and inference results closely
    together. This helps the reader, and makes the figures stand more for
    themselves.
\end{point}

\reply{
    This is a good idea,
    but we haven't been able to see how to implement in a way that would improve things.
    Maybe we don't see your idea,
    but the additional graphs that describe population structure in the Appendix
    won't be legible if added to existing figures,
    and we think we already have more than enough figures.
}

\begin{point}{}
    Useful result to know: Maryuama’s invariance principle (Nagylaki 1998):
    That the correctly weighted within coalescence time within demes only
    depends on the total population size  (given conservative migration) -- so
    it cannot fluctuate too wildly across demes and movement rates! This fact
    motivates why not fitting too wildly differing within-deme coalescence
    times/diversities.
    % Nagylaki, T., 1998 The expected number of heterozygous sites in a subdivided
    % population. Genetics 149: 1599
\end{point}

\reply{
    Good reference! We'd already referred to the simpler version of \citet{strobeck1987average},
    and have now added this reference. \revref
}

@@

\begin{point}{\revref}
    400,000 generations will look a bit lunatic for someone not in
    popgen who does not know about rescaling. You could give re-scaled, more
    realistic values!
\end{point}

\reply{
    We didn't really intend that number to sound so outlandish,
    but we think it's more ``quite long'' than ``absurdly long''?
    We're not sure that mentioning rescaling will help,
    but have added a reassuring note. {\revref} 
    Note that we \emph{did} actually simulate for that many generations;
    we didn't use rescaling to run a shorter or smaller simulation.
}

@@

\begin{point}{Paragraph beginning \revref:}
    Maybe move this section in front of the inference part, as it is useful to
    know the most ``fundamental'' limit of resistance distances.
\end{point}

\reply{
    This is a good suggestion, 
    but we'd rather keep the different results having to do with the same model
    together in the same place.
}

@@

\begin{point}{Fig. 7:}
    With EEMS one can see very suspiciously the 4x4 sampling structure - there
    are red dots in almost all 3x3 centers of the sampling squares. Since EEMS
    is quite fast, could you increase the resolution of sampling?
\end{point}

\reply{
    More samples could be added for the EEMS runs, 
    but doing so would have the drawback 
    that the two methods would no longer be using the same data for inference. 
    We've been told that similar patterns show up when EEMS tries to approximate
    anisotropic migration,
    so it may not be the sampling pattern.
    We've added a note about this. {\revref}
    Also, note that these samples are more spread out that people often have in practice:
    they do occur in the middle three-quarters of each grid square,
    but they aren't all located at the center.
}

@@

\begin{point}{}
    You could experiment with visualizations a bit. EEMS got quite popular as
    exploratory visualization tool because it produces visually pleasing plots.
    Do lines with two arrowheads of different sizes near the middle, maybe even
    colored (blue to red) work?
\end{point}

\reply{
    We agree -- there's a lot of directions we could move in to improve our package,
    and hope to work more on this.
}

@@

\reviewersection{2}

\begin{quote}
    This work revisits McRae’s seminal isolation-by-resistance (IBR) model through
    the lens of coalescent theory. The original idea of IBR was to compare patterns of
    genetic distance observed on a landscape with the so-called ``resistance distance''.
    The resistance distance between states $i$ and $j$ in a Markov chain is defined to be
    the ``commute time'': the expected amount of time it takes for the chain to go
    from $i$ to $j$ and back, integrated over all possible paths of the chain. Resistance
    distance comes from electrical circuit theory and is not a model of genetic
    variation across a landscape. Also, it is inherently symmetric in $i$ and $j$, and
    cannot accomodate asymmetric migration between demes. A more biologically
    realistic model is the structured coalescent. Hence, if data are generated under
    the something resembling the coalescent, then there is scope for improving IBR
    by reducing model misspecification.

@@

    This paper implements such a method and conducts comparative analyses on a
    number of simulated data sets. It seems that in general, this method and the
    resistance-based approaches (as exemplified by EEMS) produce somewhat com-
    parable results so long as migration rates are symmetric. The largest differences
    emerge when gene flow is asymmetric, because this fundamentally violates the
    resistance distance model used by EEMS. Here, the authors show convincingly
    that inferring migration rates using the resistance distance approximation can
    be wildly inaccurate compared to the same estimates obtained using coalescent
    methods. To the extent that gene flow is more likely to be asymmetric in many
    natural populations, this seems like an important finding.

    My overall impression of this work is fairly positive. The paper is very clearly
    written, and the idea of replacing IBR with a method that is more faithful to
    the underlying biology is appealing. The methods section (as well as appendices)
    feature a very nice and readable summary of the mathematical innovations of
    this work, as well as some interesting discussion of issues of identifiability and
    power. Similarly, discussion section in very well done and really strengthens
    the manuscript. It includes a thorough survey of the related work, as well as a
    forthright discussion of advantages and disadvantages of the proposed method,
    and it raises a number of interesting questions and directions for future research.
    Overall, the paper was a pleasure to read.
\end{quote}

Thanks very much for the kind words.
One minor comment: resistance distances are not strictly based in a population genetics model,
but we think that in practice it is thought of as (an approximation to) such a model
-- for instance, it is common to see the explanation that it
``integrates over all possible paths that the organisms might take''.
I think it is clear that \citet{mcrae2006isolation} saw resistance distance as an approximation
to the more-accurate coalescence models; we're looking at the effects of this approximation.
We hope this point comes through in the paper.

\begin{quote}
    I find the main deficiency to be the lack of any analysis of real data. All
    experiments are carried out on simulated data only. This is a puzzling omission.
    The authors went to the trouble of creating an R package (which is good), so
    clearly they intend for the method to be used in practice. Personally, I would
    hesitate to publish the results obtained from a method which had not previously
    been verified to produce reasonable answers in real data. This really feels like
    something the authors ought to have included. It is not necessary that the data
    be novel -- reanalyzing (say) the African elephant dataset in the EEMS paper
    would provide for an interesting comparison (even if it cannot be done at an
    equivalent spatial resolution) and would also go long way towards reassuring
    practitioners.
\end{quote}

We've now added analysis of a real dataset, from \citet{geraldes2014landscape}.
We're quite pleased with the results.

### 2019-04-17 isolation_by_coalescence.tex
more responses

There are a number of ways to calculate measures of dissimilarity using genetic data.
In this paper we use mean pairwise divergence, \revpoint{3}{6}
which for two genomes is calculated as the proportion of sites in the genome
at which the two genomes differ. \revpoint{3}{22}
It is more common in the literature to use $F_{ST}/(1-F_{ST})$, \revpoint{1}{7}
but the two are highly correlated for good theoretical reasons 
\citep{slatkin1991inbreeding,rousset1997genetic},
especially for SNP data.
The two genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
from the two genomes back to their most recent common ancestor at that site.
Therefore,
under an infinite-sites model with average mutation rate $\mu$ per site and per meiosis,
pairwise divergence divided by $\mu$
is an unbiased estimate of the mean time to most recent common ancestor,
averaged across the genome \citep{hudson2007variance,ralph2015empirical}.
It is natural to model this process by following the two lineages back through time
(and across space)
until they find their most recent common ancestor,
i.e., until they coalesce.
Suppose we model the geographic distribution of the species
as a collection of randomly mating populations
that exchange occasional migrants,
thus discretizing geographic space.
If the genome is selectively neutral,
then the motion of these lineages back through time
forms a Markov process, the \emph{structured coalescent} \citep{wakeley2005coalescent}.
In this framework, each lineage performs a random walk across the populations,
with movement probabilities depending on the flux of individuals between populations.
This provides the link between genetic distance and random walks.

@@

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so after rearrangement can be solved with standard linear algebra
(and is related to the Sylvester equation \citep{bhatia1997solve}).
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZC$ is skew-symmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $\epsilon$
so that $G + \epsilon Z$ remains the generator of a valid Markov chain.

@@

\paragraph{MCMC methods}
To sample from the posterior distribution of equation \eqref{eq:post}, 
we used R \citep{Rmanual} 
to implement a standard Metropolis-Hastings algorithm with Gaussian proposals \citep{brooks2011handbook}.
Starting locations were chosen by sampling from the prior distribution.
Before the usual ``burn-in'' phase,
we included a period of ``pre-burn-in'' which used the same MCMC procedure
with a larger $\sigma_\epsilon$, to allow the chain to more quickly converge on the high-posterior
portion of parameter space.
Typically, we ran MCMC for $10^6$ iterations of pre-burn-in 
and $3 \times 10^6$ iterations of burn-in,
followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions,
although for small graphs fewer iterations were necessary.
We used standard methods to assess convergence and mixing.

@@

\subsection*{Code availability}

@@

@@

>>>>

There are a number of ways to calculate measures of dissimilarity using genetic data.
In this paper we use mean pairwise divergence, \revpoint{3}{6}
which for two genomes is calculated as the proportion of sites in the genome
at which the two genomes differ. \revpoint{3}{22}
It is more common in the literature to use $F_{ST}/(1-F_{ST})$, \revpoint{1}{7}
but the two are highly correlated for good theoretical reasons 
\citep{slatkin1991inbreeding,rousset1997genetic},
especially for SNP data.
(However, note that $F_{ST}$ depends on the sampling scheme, while divergence does not.) \revpoint{1}{7}
The two genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
from the two genomes back to their most recent common ancestor at that site.
Therefore,
under an infinite-sites model with average mutation rate $\mu$ per site and per meiosis,
pairwise divergence divided by $\mu$
is an unbiased estimate of the mean time to most recent common ancestor,
averaged across the genome \citep{hudson2007variance,ralph2015empirical}.
It is natural to model this process by following the two lineages back through time
(and across space)
until they find their most recent common ancestor,
i.e., until they coalesce.
Suppose we model the geographic distribution of the species
as a collection of randomly mating populations
that exchange occasional migrants,
thus discretizing geographic space.
If the genome is selectively neutral,
then the motion of these lineages back through time
forms a Markov process, the \emph{structured coalescent} \citep{wakeley2005coalescent}.
In this framework, each lineage performs a random walk across the populations,
with movement probabilities depending on the flux of individuals between populations.
This provides the link between genetic distance and random walks.

@@

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so after rearrangement can be solved with standard linear algebra
(and is related to the Sylvester equation \citep{bhatia1997solve}).
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZC$ is skew-symmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
For $G + Z$ to be a valid generator, the rows of $Z$ must sum to zero,
and all offdiagonal entries of $G + Z$ must be nonnegative. \revpoint{1}{11}
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $\epsilon$
so that $G + \epsilon Z$ remains the generator of a valid Markov chain.

@@

\paragraph{MCMC methods}
To sample from the posterior distribution of equation \eqref{eq:post}, 
we used R \citep{Rmanual} 
to implement a standard Metropolis-Hastings algorithm with Gaussian proposals \citep{brooks2011handbook}.
Starting locations were chosen by sampling from the prior distribution.
Before the usual ``burn-in'' phase,
we included a period of ``pre-burn-in'' which used the same MCMC procedure
with a larger $\sigma_\epsilon$, to allow the chain to more quickly converge on the high-posterior
portion of parameter space.
Typically, we ran MCMC for $10^6$ iterations of pre-burn-in 
and $3 \times 10^6$ iterations of burn-in,
followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions,
which took around 10 hours on a single core of a modern computer with 16 demes. \revpoint{1}{1}
For small graphs fewer iterations were necessary.
We used standard methods to assess convergence and mixing. \revpoint{1}{6}

@@

%%%%%%%%%%%%%%%
\subsection*{Data from \textit{Populus trichocarpa}}

We also applied our method to genotyping data from 423 samples of \textit{Populus trichocarpa}
\citet{geraldes2014landscape}
and 8 samples of the closely related \textit{Populus balsamifera} (with which it hybridizes),
from a large region of northwestern North America,
described in \citet{geraldes2014landscape}.
These data \citep{geraldes2014data},
include geographical coordinates and genotypes from a genotyping chip targeting 34K SNPs.
In addition to preliminary filtering described in \citet{geraldes2014landscape},
we additionally removed 2,314 SNPs with more than 5\% missing data, retaining 30,756 SNPs,
and two individuals, one with more than 15\% missing data and one that is much further east
than all the other samples.
We then computed pairwise divergence between the remaining samples.
We discretized the landscape manually into nine regions,
preserving species boundaries and choosing divisions along gaps in sampling
to keep geographic areas roughly equal.
The numbers of samples included in each discretized region varied from 1 to 268,
but our method does not assume equal sample numbers
(however, an $F_{ST}$-based method would likely be affected by such large differences in sampling).

%%%%%%%%%%%%%%%%
\subsection*{Code availability}

@@

%%%%%%%%%%%%%
\subsection*{Application to \textit{Populus trichocarpa}}

Our discretization of the sampling area for the two poplar species
is shown in Figure~\ref{fig:populus_results},
along with

@@

Computation time currently limits this method to a discretization of less than 30 populations,
but nonidentifiability is probably a more serious barrier that must be concurrently addressed
before scaling the method to finer discretizations of space. \llabel{ll:computation}
Finer discretizations of space should benefit from the use of sparse matrix methods,
but in our testing these did not speed up computation at this scale. \revpoint{1}{2}
